\subsection{Navier-Stokes Equations} 

\subsubsection{Governing equations:}

The Navier-Stokes equations arise from applying Newton's second law to fluid
motion, \ie the temporal and spatial fluid inertia is in equilibrium with
internal (volume/body) and external (surface) forces. The reaction of surface
forces can be described in terms of the fluid stress as the sum of a diffusing
viscous term, plus a pressure term. The equations are named for the 19th
century contributions of Claude-Louis Navier and George Gabriel Stokes. A
solution of the Navier-Stokes equations is called a flow field, \ie velocity
and pressure field, which is a description of the fluid at a given point in
space and time.  In the common case of an incompressible Newtonian fluid, the
nonlinear Navier-Stokes equations (three-dimensional, transient) can be
written using 'primitive variables' (\ie $u$-velocity, $p$-pressure) as:
\begin{equation}
    \rho\delby{\vectr{u}}{t}+\rho\pbrac{\dotprod{\vectr{u}}{\grad}}\vectr{u}=\vectr{f}-\gradient{}{p}+\mu\laplacian{}{\vectr{u}}
  \label{eqn:NSE}
\end{equation}
accompanied by the conservation of mass (incompressibility)
\begin{equation}
  \divergence{}{\vectr{u}}=0
  \label{eqn:NSEIncompressibility}
\end{equation}
where $\vectr{u}(\vectr{x},t)=\transpose{\pbrac{u_{1},u_{2},u_{3}}}$ is the
velocity vector depending on spatial coordinates
$\vectr{x}=\transpose{\pbrac{x_{1},x_{2},x_{3}}}$ and the time $t$, $p$ is the
scalar pressure, $\vectr{f}$ an applied body force, and the material
parameters $\mu$ and $\rho$ are the fluid viscosity and density,
respectively.The first term represents unsteady accelerative inertial
contributions, the second represents the nonlinear convective acceleration
terms, the $\gradient{}{p}$ term the pressure contributions, and the last term
represents viscous stresses in the system.
% incompressibility and the LBB condition
As with Stokes flow, the incompressibility condition
$\divergence{}{\vectr{u}}=0$ also creates restrictions on the formulation of
the velocity and pressure spaces using finite elements known as the
Ladyzhenkaya, Babuska, and Brezzi (LBB) or inf-sup consistency
condition. Several methods have been devised to define a pressure function
that is consistent with the velocity space using primitive variables. These
include mixed element methods, penalty methods, generalized petrov-galerkin
methods using pressure poisson correction, operator splitting, and
semi-implicit pressure correction \cite{chung:2010}.

Using a classic Galerkin formulation, mixed methods are perhaps conceptually
the most straightforward method of satisfying LBB, in which velocity is
defined over a space one order higher than pressure (e.g. quadratic elements
for velocity, linear for pressure), allowing incompressibility to be
satisfied. It should be noted that our use of a mixed formulation to satisfy
LBB will also be reflected in the shape functions that our weak formulation
depends on. For example, using a 2D element with biquadratic velocity and
linear pressure, we will have 9 DOFs and weight functions for each velocity
component and 4 for the pressure.
% meaning of pressure in NSE.

\subsubsection{Weak Formulation}
% Galerkin formulation
The weak form of equations \ref{eqn:NSE} and \ref{eqn:NSEIncompressibility}
can be found by applying standard Galerkin test functions $\vectr{w}$:
\begin{equation}
  \gint{\Omega}{}{\pbrac{\rho\delby{\vectr{u}}{t}
    +\rho\vectr{u}\cdot\gradient{}{\vectr{u}}
    -\vectr{f}
    -\mu\laplacian{}{\vectr{u}}
    +\gradient{}{p}}{\vectr{w}}}{\Omega} = 0
 \label{eqn:BasicGalerkinNSE}
\end{equation}

Integrating by parts, we will get the weak form of the system of PDEs with the associated natural boundary conditions at the boundary $\Gamma_N$:
\begin{multline}
  \gint{\Omega}{}{\rho\delby{\vectr{u}}{t}{\vectr{w}}}{\Omega}
 +\gint{\Omega}{}{\rho\vectr{u}\cdot\grad\vectr{u}{\vectr{w}}}{\Omega}
 +\gint{\Omega}{}{p\gradient{}{\vectr{w}}}{\Omega}
 -\gint{\Omega}{}{\mu\gradient{}{\vectr{u}}\cdot\vectr{\gradient{}{w}}}{\Omega}=\\
 \gint{\Omega}{}{\vectr{f}\vectr{w}}{\Omega}
 +\gint{\Gamma_N}{}{\mu\dotprod{\grad\vectr{u}}{\vectr{n}}\vectr{w}}{\Gamma_N}
 -\gint{\Gamma_N}{}{\dotprod{p}{\vectr{n}}\vectr{w}}{\Gamma_N}
 \label{GalerkinNSE}
\end{multline}

% Boundary conditions
For more extensive discussion of this procedure, along with other weak forms
of the PDEs, we refer to \cite{gresho:2000}. From this weak form, we see
natural (Neumann) boundary conditions arising as a direct result of the
integration. Neumann boundary conditions will consist of a pressure term and
viscous stress acting normal to a given boundary.
\begin{gather}
   \fnof{\vectr{q}}{\vectr{x},t} = \mu\dotprod{\gradient{}{\fnof{\vectr{u}}{\vectr{x},t}}}{\normal}-p\cdot{\normal}  \\
%   \fnof{\vectr{q_2}}{\vectr{x},t} = p\cdot{\normal} \\
   \quad \vectr{x}\in\Gamma_{N}
 \label{eqn:NSENeumannBC}  
\end{gather}
Specification of Neumann boundaries will simply require the specification of
the terms across element DOFs. Dirichlet boundary conditions on a boundary
$\Gamma_D$ for velocity will take the form:
\begin{gather}
  \fnof{\vectr{u}}{\vectr{x},t} = \fnof{\vectr{d}}{\vectr{x},t} \quad \vectr{x}\in\Gamma_{D}\\
 \label{eqn:NSEDirichletBC} 
\end{gather}

\subsubsection{Tensor notation}
Assuming no forcing terms and substituting the natural boundary as defined
above, equation \ref{eqn:GalerkinNSE} in tensor notation becomes:
\begin{multline}
  \gint{\Omega}{}{\rho\dot{u}_{i}w_{i}}{\Omega}
 +\gint{\Omega}{}{{\rho}G^{jk}u_{j}\covarderiv{u_{i}}{k}w_{i}}{\Omega}
 +\gint{\Omega}{}{G^{jk}{p_{i}}\covarderiv{w_{i}}{k}}{\Omega}\\
 -\gint{\Omega}{}{{\mu}G^{jk}\covarderiv{u_{i}}{j}\covarderiv{w_{i}}{k}}{\Omega}
 -\gint{\Gamma_N}{}{{q_{i}}{w_i}}{\Gamma_N}=\vectr{0}
 \label{eqn:TensorNSE}
\end{multline}
or
\begin{multline}
  \gint{\Omega}{}{\rho\dot{u}_{i}w_{i}}{\Omega}
 +\gint{\Omega}{}{{\rho}G^{jk}{u_j}\pbrac{\partialderiv{u_{i}}{k}-\christoffel{i}{k}{h}u_h}{w_{i}}}{\Omega}
 +\gint{\Omega}{}{G^{jk}{p_{i}}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}\\
 -\gint{\Omega}{}{{\mu}G^{jk}\pbrac{\partialderiv{u_{i}}{j}-\christoffel{i}{j}{h}u_h}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}
 -\gint{\Gamma_N}{}{{q_{i}}{w_i}}{\Gamma_N}=\vectr{0}
 \label{eqn:Tensor2NSE}
\end{multline}

where $G^{jk}$ is the contravariant metric tensor and $\christoffel{i}{j}{k}$
is the Christoffel symbol of the second kind for the spatial coordinates.

\subsubsection{Finite Element Formulation}

We can now discretise the domain into finite elements \ie
$\Omega=\displaystyle{\bigcup_{e=1}^{E}}\Omega_{e}$ with
$\Gamma=\displaystyle{\bigcup_{f=1}^{F}}\Gamma_{f}$. \Eqnref{eqn:Tensor2NSE} now
becomes:

\begin{multline}
  \dsum_{e=1}^{E}\gint{\Omega}{}{\rho\dot{u}_{i}w_{i}}{\Omega}
 +\dsum_{e=1}^{E}\gint{\Omega}{}{{\rho}G^{jk}{u_j}\pbrac{\partialderiv{u_{i}}{k}-\christoffel{i}{k}{h}u_h}{w_{i}}}{\Omega}
 +\dsum_{e=1}^{E}\gint{\Omega}{}{G^{jk}{p_{i}}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}\\
 -\dsum_{e=1}^{E}\gint{\Omega}{}{{\mu}G^{jk}\pbrac{\partialderiv{u_{i}}{j}-\christoffel{i}{j}{h}u_h}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}
 -\dsum_{f=1}^{F}\gint{\Gamma_N}{}{{q_{i}}{w_i}}{\Gamma_N}=\vectr{0}
 \label{eqn:FEMNSE}
\end{multline} 
or
\begin{multline}
  \dsum_{e=1}^{E}\gint{\Omega}{}{\rho\dot{u}_{i}w_{i}}{\Omega}
 +\dsum_{e=1}^{E}\gint{\Omega}{}{{\rho}G^{jk}{u_j}\pbrac{\partialderiv{u_{i}}{k}-\christoffel{i}{k}{h}u_h}{w_{i}}}{\Omega}
 +\dsum_{e=1}^{E}\gint{\Omega}{}{G^{jk}{p_{i}}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}\\
 -\dsum_{e=1}^{E}\gint{\Omega}{}{{\mu}G^{jk}\pbrac{\partialderiv{u_{i}}{j}-\christoffel{i}{j}{h}u_h}\pbrac{\partialderiv{w_{i}}{k}-\christoffel{i}{k}{h}w_h}}{\Omega}=\\
 \dsum_{f=1}^{F}\gint{\Gamma_N}{}{{\mu}G^{jk}\pbrac{\partialderiv{u_{i}}{k}-\christoffel{i}{k}{h}u_h}{n_{i}}{w_{i}}}{\Gamma_N}
 -\dsum_{f=1}^{F}\gint{\Gamma_N}{}{G^{jk}{p}{n_i}{w_i}}{\Gamma_N}
 \label{eqn:FEMNSE2}
\end{multline} 

We will assume that the dependent variables $\vectr{u}$ and $p$ can be
interpolated separately in space and time. Here we must also be careful to
note again the discrepancy between the functional spaces for velocity and
pressure using a mixed formulation to satisfy the LBB consistency
requirement. We will therefore define two separate weighting functions: for
the velocity space on $\Omega$, the test function will be $\psi_i$ and for the
pressure space, $\phi_i$, giving:

\begin{gather}
  \fnof{\vectr{u}}{\vectr{x},t}=\gbfn{n}{}{\vectr{x}}\fnof{\nodept{\vectr{u}}{n}}{t}\\
  \fnof{p}{\vectr{x},t}=\phi_{n}({\vectr{x}})\fnof{\nodept{p}{n}}{t}
\end{gather}

In standard interpolation notation within an element, we transform from
$\vectr{x}$ to $\vectr{\xi}$:

\begin{gather}
  \fnof{u_{i}}{\vectr{\xi},t}=\idxgbfn{i}{n}{\beta}{\vectr{\xi}}
  \fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\\
  \fnof{p_{}}{\vectr{\xi},t}=\phi_{in}^{\beta}({\vectr{\xi}})
  \fnof{\idxnodedof{p}{}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}
\end{gather}

where $\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}$ are the time varying nodal
degrees-of-freedom for velocity component $i$, node $n$, global derivative
$\beta$, $\idxgbfn{i}{n}{\beta}{\vectr{\xi}}$ are the corresponding velocity
basis functions and $\idxgsf{i}{n}{\beta}$ are the scale factors. The scalar
pressure DOFs, $\fnof{\idxnodedof{p}{}{n}{\beta}}{t}$ are interpolated
similarly.

For the natural boundary, we can separate $q_i$ into its component velocity
and pressure terms as noted in \ref{eqn:NSENeumannBC} and shown in
\ref{eqn:FEMNSE2}. For our current treatment, we will also assume the values
of viscosity $\mu$ and density $\rho$ are constant. These can be interpolated:

\begin{equation}
  \begin{split}
    \fnof{q_{i}}{\vectr{\xi},t} &= \idxgbfn{i}{o}{\gamma}{\vectr{\xi}}
      \fnof{\idxnodedof{q}{i}{o}{\gamma}}{t}\idxgsf{i}{o}{\gamma} \\
    \fnof{\mu}{\vectr{\xi}} &=\gbfn{r}{\delta}{\vectr{\xi}}
    \nodedof{\mu}{r}{\delta}\gsf{r}{\delta} \\
    \fnof{\rho}{\vectr{\xi}} &=\gbfn{r}{\delta}{\vectr{\xi}}\nodedof{\rho}{r}{\delta}
    \gsf{r}{\delta} \\
  \end{split}
\end{equation}

Using the spatial weighting functions for a Galerkin finite element
formulation:
\begin{equation}
  \fnof{w_{i}}{\vectr{\xi}}=\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
\end{equation}


\subsubsection{Spatial Integration}

Using standard integration notation, we can rewrite our Galerkin FEM
formulation from \ref{eqn:NSEFEM2}:

\begin{multline}
%time dependence
  \dsum_{e=1}^{E}\gint{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}
      \fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}}{t}
    \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}\\
%convective term
  - \dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}}G^{jk}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
  \fnof{\idxnodedof{u}{j}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\\
  \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}
      \idxgsf{i}{n}{\beta}}{x^{k}}-\christoffel{i}{k}{h}\idxgbfn{h}{m}{\alpha}{\vectr{\xi}}
    \fnof{\idxnodedof{u}{h}{n}{\beta}}{t}\idxgsf{h}{m}{\alpha}}
  \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%pressure
  +\dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}G^{jk}\phi_{in}^{\beta}({\vectr{\xi}})
  \fnof{\idxnodedof{p}{}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\\
  \pbrac{\delby{\psi_{im}^{\alpha}({\vectr{\xi}})\idxgsf{i}{m}{\alpha}}{x^{k}}-
  \christoffel{i}{k}{h}\psi_{im}^{\alpha}({\vectr{\xi}})\idxgsf{h}{m}{\alpha}
    }\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%viscous stress
  -\dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}{\fnof{\mu}{\vectr{\xi}}}G^{jk}\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}
     \fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}}{x^{j}}-
      \christoffel{i}{j}{h}\idxgbfn{h}{n}{\beta}{\vectr{\xi}}
      \fnof{\idxnodedof{u}{h}{n}{\beta}}{t}\idxgsf{h}{n}{\beta}} \\
    \pbrac{\delby{\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}}{x^{k}}-
      \christoffel{i}{k}{l}\idxgbfn{l}{m}{\alpha}{\vectr{\xi}}
      \fnof{\idxnodedof{u}{h}{n}{\beta}}{t}
      \idxgsf{l}{m}{\alpha}}\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%boundary terms
  =\dsum_{f=1}^{F}\gint{\vectr{0}}{\vectr{1}}{\idxgbfn{i}{o}{\gamma}{\vectr{\xi}}
    \fnof{\idxnodedof{q}{i}{o}{\gamma}}{t}\idxgsf{i}{o}{\gamma}
    \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}
  \label{eqn:NSEFEM3}
% note: alternate symbol for normal vector needed if want to expand boundary terms
%=\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}
%      \idxgsf{i}{n}{\beta}}{x^{k}}-\christoffel{i}{k}{h}\idxgbfn{h}{m}{\alpha}{\vectr{\xi}}
%    \fnof{\idxnodedof{u}{h}{n}{\beta}}{t}\idxgsf{h}{m}{\alpha}}\\
%  \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
%  \abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
\end{multline}

Where $\fnof{\matr{J}}{\vectr{\xi}}$ represents the jacobian matrix. If we
assume a rectangular cartesian coordinate system, this simplifies
significantly, as the contravariant $G^{jk}$ will become $\delta^{jk}$ and the
Christoffel symbols $\christoffel{i}{j}{k}$ will all be zero. This gives:

\begin{multline}
 \label{NSEFEM3}
%time dependence
  \dsum_{e=1}^{E}\gint{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}\delby{(\idxgbfn{i}{n}{\beta}{\vectr{\xi}}
      \fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta})}{t}
    \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}\\
%convective term
  - \dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}}\delta^{jk}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
  \fnof{\idxnodedof{u}{j}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}
  \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}
   \idxgsf{i}{n}{\beta}}{x^{k}}}\\
  \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%pressure
  +\dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}\delta^{jk}\phi_{in}^{\beta}({\vectr{\xi}})
  \fnof{\idxnodedof{p}{}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}
  \pbrac{\delby{\psi_{im}^{\alpha}({\vectr{\xi}})\idxgsf{i}{m}{\alpha}}{x^{k}}}\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%viscous stress
  -\dsum_{e=1}^{E}\dintl{\vectr{0}}{\vectr{1}}{\fnof{\mu}{\vectr{\xi}}}\delta^{jk}\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}
     \fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}}{x^{j}}}
    \pbrac{\delby{\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}}{x^{k}}}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
%boundary terms
  =\dsum_{f=1}^{F}\gint{\vectr{0}}{\vectr{1}}{\idxgbfn{i}{o}{\gamma}{\vectr{\xi}}
    \fnof{\idxnodedof{q}{i}{o}{\gamma}}{t}\idxgsf{i}{o}{\gamma}
    \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}\idxgsf{i}{m}{\alpha}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}
\end{multline} 

or, rearranged to pull constants out of the integrals:

\begin{multline}
 \label{NSEFEM4}
%time dependence
  \dsum_{e=1}^{E}{\idxnodedof{\dot{u}}{i}{n}{\beta}}(t){\idxgsf{i}{n}{\beta}}{\idxgsf{i}{m}{\alpha}}
  \gint{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}}{\vectr{\xi}}\\
 %convective term
  -\dsum_{e=1}^{E}{\idxnodedof{u}{j}{n}{\beta}}(t){\idxgsf{i}{n}{\beta}}^2{\idxgsf{i}{m}{\alpha}}
   \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}{\fnof{\rho}{\vectr{\xi}}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
   \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
     \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{k}}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}d{\vectr{\xi}}\\
 %pressure
    +\dsum_{e=1}^{E}{\fnof{\idxnodedof{p}{}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\idxgsf{i}{m}{\alpha}}
    \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}\phi_{in}^{\beta}({\vectr{\xi}})
    \pbrac{\delby{\psi_{im}^{\alpha}({\vectr{\xi}})}{x^{k}}}\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
 % %viscous stress
    -\dsum_{e=1}^{E}\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\idxgsf{i}{m}{\alpha}
    \dintl{\vectr{0}}{\vectr{1}}{\delta^{jk}\fnof{\mu}{\vectr{\xi}}}\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{j}}}
      \pbrac{\delby{\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}}{x^{k}}}
      \abs{\fnof{\matr{J}}{\vectr{\xi}}}d{\vectr{\xi}}\\
% %boundary terms
  =\dsum_{f=1}^{F}\fnof{\idxnodedof{q}{i}{o}{\gamma}}{t}\idxgsf{i}{m}{\alpha}\idxgsf{i}{o}{\gamma}
   \gint{\vectr{0}}{\vectr{1}}{\gbfn{m}{\alpha}{\vectr{\xi}}\gbfn{o}{\gamma}{\vectr{\xi}}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}
\end{multline}


\subsubsection{General form}

We now seek to assemble this into the corresponding general form for dynamic equations, as outlined in \secref{sec:dynamicequations}:

\begin{equation}
  \matr{M}\fnof{\ddot{\vectr{d}}}{t}+\matr{C}\fnof{\dot{\vectr{d}}}{t}+\matr{K}\fnof{\vectr{d}}{t}+
  \fnof{\vectr{g}}{\fnof{\vectr{d}}{t}}+\fnof{\vectr{f}}{t}=\vectr{0}
  \label{eqn:NSEGeneralDynamic}
\end{equation}
where $\dot{\vectr{d}}(t)$ and $\ddot{\vectr{d}}(t)$ represent the first and
second derivatives (respectively) of the degrees of freedom vector
$\vectr{d}(t)$, which consists of the dependent variables
$\vectr{u}(\vectr{x},t)$ and $p(x,t)$. $\matr{M}$ is the mass matrix, which
provides the shape function based weights and $\matr{C}$ is the transient
damping matrix. $\matr{K}$ represents the stiffness matrix, which will contain
the linear parts of the operator. $\fnof{\vectr{g}}{\fnof{\vectr{u}}{t}}$ is
the nonlinear vector function that will be used to represent the convective
term and $\fnof{\vectr{f}}{t}$ the forcing vector.

We will assume cartesian coordinates $\vectr{x}=\bbrac{x,y,z}$ and denote the
corresponding velocity components $\vectr{u}=\bbrac{u,v,w}$, with $N$
representing the number of velocity DOFs and $M$ the number of pressure
DOFs. The vector $d$ in then becomes:
\begin{equation}
  \vectr{d}=[u_1u_2...u_Nv_1v_2...v_Nw_1w_2...w_Np_1p_2...p_M]^T=
  \begin{bmatrix}\vectr{u}\\\vectr{v}\\\vectr{w}\\\vectr{p}\end{bmatrix}
\end{equation}

All the components of \ref{eqn:NSEGeneralDynamic} will similarly depend on the
number of dimensions,$n_{dim}$, and the size of the matrices will be:
$(N{\cdot}{n_{dim}}+M)^{n_{dim}}$.

Returning to the general case, the mass matrix $\matr{M}$ will take the form:

\begin{equation}
[M^{\alpha\beta}_{mn}]=\gint{\vectr{0}}{\vectr{1}}{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}
\end{equation}
Without using mass-lumping, the damping matrix $\matr{C}$:
\begin{equation}
  [C^{\alpha\beta}_{mn}]={\idxgsf{i}{n}{\beta}}{\idxgsf{i}{m}{\alpha}}
  \gint{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}}{\vectr{\xi}}
\end{equation}

The element stiffness matrix $\matr{K}$ will contain the linear components of
the system. For the classic Galerkin formulation this includes the viscous
term with the laplacian operator and the pressure gradient term. As the
velocity and pressure DOFs are both included in $d$, $\matr{K}$ will be
assembled so that the corresponding operators are applied to the DOFs
discontinuously.

\begin{equation}
 [K^{\alpha\beta}_{mn}]=
 \begin{cases}
   [A^{\alpha\beta}_{mn}] =  {\idxgsf{i}{n}{\beta}{\idxgsf{i}{m}{\alpha}}}
    \dintl{\vectr{0}}{\vectr{1}}{\delta^{jk}\fnof{\mu}{\vectr{\xi}}}\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{j}}}
       \pbrac{\delby{\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}}{x^{k}}}
       \abs{\fnof{\matr{J}}{\vectr{\xi}}}d{\vectr{\xi}}\\
     \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad \text{for $n\le{N{\cdot}{N_{dim}}}$},\\

   [B^{\alpha\beta}_{mn}] = \idxgsf{i}{n}{\beta}\idxgsf{i}{m}{\alpha}
     \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}\phi_{in}^{\beta}({\vectr{\xi}})
     \pbrac{\delby{\psi_{im}^{\alpha}({\vectr{\xi}})}{x^{k}}}\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
    \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad \text{for $n>{N{\cdot}{N_{dim}}}$}.
 \end{cases}
\end{equation}

To attain a square matrix of the required size, we must also apply the
transpose of the gradient terms acting on the pressure DOFs. For example, in a
3 dimensional example system, $\matr{K}$ will take the form:
\begin{equation}
  \matr{K}=
   \begin{bmatrix}
     \matr{A} & \matr{0} & \matr{0} & \matr{B}_x\\
     \matr{0} & \matr{A} & \matr{0} & \matr{B}_y\\
     \matr{0} & \matr{0} & \matr{A} & \matr{B}_z\\
     -\matr{B}^T_{x}& -\matr{B}^T_{y} & -\matr{B}^T_z & \matr{0}
   \end{bmatrix}
\end{equation}

The nonlinear vector, $\vectr{g}(t)$ will provide the convective operators and
for the standard Galerkin formulation:

\begin{equation}
 [g^{\alpha\beta}_{mn}]=
 {\idxgsf{i}{n}{\beta}}^2{\idxgsf{i}{m}{\alpha}}
   \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}{\fnof{\rho}{\vectr{\xi}}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
   \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
     \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{k}}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}d{\vectr{\xi}}
\end{equation}

\subsubsection{Streamline Upwind/Petrov-Galerkin (SUPG)}

For convection dominated flows, it is often useful to modify the Galerkin
formulation account for numerical problems associated with large asymmetric
velocity gradients. We describe augmenting the Galerkin Navier-Stokes
formulation to form a Petrov-Galerkin form, which uses a different test
function for the convective term to stabilize the algorithm with a balancing
diffusive term. Please refer to \secref{sec:Convective Transport} for
more details.

For the Navier-Stokes equations (and most nonlinear problems), the use of a
Petrov-Galerkin formulation becomes more difficult than for linear cases like
advection-diffusion, as consistent weighting can cause instabilities when used
with additional terms like body forces \cite{heinrich:1999}. These issues can
be overcome with more complex Petrov-Galerkin formulations, as those derived
by Hughes and colleagues in the 1980s. A useful alternative route is to apply
SUPG weights to the convective operator in elements as a function of each
element's Reynolds number (referred to subsequently as the cell Reynolds
number). This can provide the desired stabilizing effect in the direction of
large velocity gradients but can be easily implemented and is practically
useful.

Assuming cartesian coordinates and applying Petrov-Galerkin weights to the
convective term, the finite element formulation described in equation
\ref{eqn:NSEFEM2} can be written:

\begin{multline}
  \dsum_{e=1}^{E}\gint{\Omega}{}{\rho\dot{u}_{i}w_{i}}{\Omega}
  +\dsum_{e=1}^{E}\gint{\Omega}{}{{\rho}{\delta}^{jk}{u_j}\pbrac{\partialderiv{u_{i}}{k}}({w_{i}+\Psi_{i}})}{\Omega}
  +\dsum_{e=1}^{E}\gint{\Omega}{}{{\delta}^{jk}{p_{i}}{\partialderiv{w_{i}}{k}}}{\Omega}\\
  -\dsum_{e=1}^{E}\gint{\Omega}{}{{\mu}{\delta}^{jk}\pbrac{\partialderiv{u_{i}}{j}}\pbrac{\partialderiv{w_{i}}{k}}}{\Omega}=
  \dsum_{f=1}^{F}\gint{\Gamma_N}{}{{\mu}{\delta}^{jk}\pbrac{\partialderiv{u_{i}}{k}}{n_{i}}{w_{i}}}{\Gamma_N}
  -\dsum_{f=1}^{F}\gint{\Gamma_N}{}{{\delta}^{jk}{p}{n_i}{w_i}}{\Gamma_N}
  \label{eqn:NSEPGFEM1}
\end{multline}

where $w_i$ will be the classic Galerkin weight and $\Psi_i$ will be the
Petrov-Galerkin for the Navier-Stokes equations. We must also define the
Peclet number, $\gamma$, for the Navier-Stokes equations to describe the ratio
of the convective:viscous operators. Here, the effective Peclet number will be
based on the cell Reynolds number:

\begin{equation}
  Re = \frac{\rho\bar{U}L}{\mu}  =\frac{\bar{U}L}{\nu}
\end{equation}
where $\bar{U}$ is the characteristic velocity and $\nu=\frac{\mu}{\rho}$ is
the kinematic viscosity. $L$ will be the characteristic length scale for a
given element. The effective Peclet number $\gamma$ will be:

\begin{equation}
  \gamma= \frac{Re}{2}  = \frac{\bar{u}L}{2\nu}
\end{equation}

We will retain the same value for
$\alpha={\coth{\frac{\gamma}{2}} - \frac{2}{\gamma}}$ as found for the
linearized advection-diffusion equation.

\begin{equation}
 \label{PGTest}
  \Psi_i = \pbrac{\frac{\alpha{L}}{2}{\frac{u_j}{\abs{u_j}}}}{\partialderiv{w_{i}}{k}}
\end{equation},

\Eqnref{eqn:NSEPGFEM1} then becomes after integration and simplification:

\begin{multline}
%time dependence
  \dsum_{e=1}^{E}{\idxnodedof{\dot{u}}{i}{n}{\beta}}(t){\idxgsf{i}{n}{\beta}}{\idxgsf{i}{m}{\alpha}}
  \gint{\vectr{0}}{\vectr{1}}{\fnof{\rho}{\vectr{\xi}}{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}}{\vectr{\xi}}\\
 %convective term
  -\dsum_{e=1}^{E}{\idxnodedof{u}{j}{n}{\beta}}(t){\idxgsf{i}{n}{\beta}}^2{\idxgsf{i}{m}{\alpha}}
   \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}{\fnof{\rho}{\vectr{\xi}}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
    \idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
     \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{k}}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}d{\vectr{\xi}}\\
 %P-G term
  -\dsum_{e=1}^{E}{\idxnodedof{u}{j}{n}{\beta}}(t){\idxgsf{i}{n}{\beta}}^2{\idxgsf{i}{m}{\alpha}}
   \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}{\fnof{\rho}{\vectr{\xi}}\idxgbfn{j}{n}{\beta}{\vectr{\xi}}
    \pbrac{\frac{(\alpha){L(\vectr{\xi})}}{2}{\frac{u_j}{\abs{u_j}}}}\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}
     \pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{k}}}
  \abs{\fnof{\matr{J}}{\vectr{\xi}}}}d{\vectr{\xi}}\\
 %pressure
    +\dsum_{e=1}^{E}{\fnof{\idxnodedof{p}{}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\idxgsf{i}{m}{\alpha}}
    \dintl{\vectr{0}}{\vectr{1}}\delta^{jk}\phi_{in}^{\beta}({\vectr{\xi}})
    \pbrac{\delby{\psi_{im}^{\alpha}({\vectr{\xi}})}{x^{k}}}\abs{\fnof{\matr{J}}{\vectr{\xi}}}d\vectr{\xi}\\
 % %viscous stress
    -\dsum_{e=1}^{E}\fnof{\idxnodedof{u}{i}{n}{\beta}}{t}\idxgsf{i}{n}{\beta}\idxgsf{i}{m}{\alpha}
    \dintl{\vectr{0}}{\vectr{1}}{\delta^{jk}\fnof{\mu}{\vectr{\xi}}}\pbrac{\delby{\idxgbfn{i}{n}{\beta}{\vectr{\xi}}}{x^{j}}}
      \pbrac{\delby{\idxgbfn{i}{m}{\alpha}{\vectr{\xi}}}{x^{k}}}
      \abs{\fnof{\matr{J}}{\vectr{\xi}}}d{\vectr{\xi}}\\
% %boundary terms
  =\dsum_{f=1}^{F}\fnof{\idxnodedof{q}{i}{o}{\gamma}}{t}\idxgsf{i}{m}{\alpha}\idxgsf{i}{o}{\gamma}
   \gint{\vectr{0}}{\vectr{1}}{\gbfn{m}{\alpha}{\vectr{\xi}}\gbfn{o}{\gamma}{\vectr{\xi}}
    \abs{\fnof{\matr{J}}{\vectr{\xi}}}}{\vectr{\xi}}\\
  \label{eqn:NSEFEM4}
\end{multline}

Notice the integration by parts is done only to the standard Galerkin weights-
this is because the artificial diffusion added by the Petrov-Galerkin terms
should only contribute within the elements

The use of the Petrov-Galerkin formulation in this way allows for the
stabilization of moderately convective problems. However, it also results in
nonsymmetric mass matrices with the additional term, making classical
mass-lumping more difficult. As a result, the use of explicit methods also
becomes more difficult for time-dependent problems.


\subsubsection{Arbitrary Lagrangian-Eulerian (ALE) Formulation}
%ALE form
Whereas \eqnref{eqn:NavierStokesequation1} has been formulated in Eulerian
form, moving domain approaches often require the ALE modification taking an
additional term into account, depending on the fluid density $\rho$ and a
correction velocity $\vectr{u}^*$ which yields to:
\begin{equation}
  \rho\pbrac{\dotprod{\pbrac{\vectr{u}-\vectr{u}^*}}{\grad}}\vectr{u}=\vectr{f}-\gradient{}{p}+\mu\laplacian{}{\vectr{u}}
  \label{eqn:NavierStokesequationALE}
\end{equation}
So far, the nonlinear term in \eqnref{eqn:NavierStokesequation1} represents
the fluid spatial acceleration only. \eqnref{eqn:NavierStokesequation2} now
also takes the dynamic inertia terms into account
\begin{equation}
  \rho\delby{\vectr{u}}{t}+ \dotprod{\rho\pbrac{\vectr{u}-\vectr{u}^*}}{\grad}\vectr{u}=\vectr{f}-\gradient{}{p}+\mu\laplacian{}{\vectr{u}}
  \label{eqn:NavierStokesequation2}
\end{equation}
which gives us the complete Navier-Stokes equations in ALE formulation.  The
following section, however, describes the reordered quasi-static formulation
of \eqnref{eqn:NavierStokesequationALE}:
\begin{equation}
  \rho\pbrac{\dotprod{\pbrac{\vectr{u}-\vectr{u}^*}}{\grad}}\vectr{u}-\mu\laplacian{}{\vectr{u}}+\gradient{}{p}=\vectr{f}
  % -\gradient{}{p}+\mu\laplacian{}{\vectr{u}}-\rho\pbrac{\dotprod{\vectr{u}^*}{\grad}\vectr{u}=\vectr{f}
  \label{eqn:NavierStokesequationALE2}
\end{equation}

\paragraph{Weak formulation:}

The corresponding weak form of the equation system consisting of
\eqnref{eqn:NavierStokesequation1} and \eqnref{eqn:NavierStokesmasequation}
can be written in the general dynamic form (see \secref{sec:dynamicequation})
\begin{equation}
  \matr{M}\fnof{\ddot{\vectr{u}}}{t}+\matr{C}\fnof{\dot{\vectr{u}}}{t}+\matr{K}\fnof{\vectr{u}}{t}+
  \fnof{\vectr{g}}{\fnof{\vectr{u}}{t}}+\fnof{\vectr{f}}{t}=\vectr{0}
  \label{eqn:generaldynamicnonlinear}
\end{equation}
where $\fnof{u}{t}$ is the dependent variables vector $\vectr{u}(\vectr{x},t)$ and $p$
for the degrees of freedom. $\matr{M}$ is the mass matrix, which provides the
shape function based weights, $\matr{C}$ is the transient damping matrix
(which we will discuss further below). $\matr{K}$ represents the stiffness
matrix, which will contain the linear parts of the operator, including the
viscous terms, the conservation of mass terms, and pressure
terms. $\fnof{\vectr{g}}{\fnof{\vectr{u}}{t}}$ is the nonlinear vector
function for the convective terms and $\fnof{\vectr{f}}{t}$ the forcing
vector.

The corresponding weak form of the equation system consisting of
\eqnref{eqn:NavierStokesequation1} and \eqnref{eqn:NavierStokesmasequation}
can be written as:
\begin{equation}
  \gint{\Omega}{}{\rho\pbrac{\dotprod{\vectr{u}}{\grad}}\vectr{u}\vectr{v} }{\Omega}
  -\gint{\Omega}{}{\rho\pbrac{\dotprod{\vectr{u}^*}{\grad}}\vectr{u}\vectr{v} }{\Omega}
  +\gint{\Omega}{}{\mu\laplacian{}{\vectr{u}}\vectr{v}}{\Omega}
  -\gint{\Omega}{}{\gradient{}{p}\vectr{v}}{\Omega}
  +\gint{\Omega}{}{\divergence{}{\vectr{u}}q}{\Omega}=
  \gint{\Omega}{}{\vectr{f}\vectr{v}}{\Omega}  
  \label{eqn:NavierStokesweakform}
\end{equation}
The general form for this kind of equation system is
\begin{equation}
  \matr{K}{\vect{\hat{u}}}+
  \fnof{\vect{\hat{g}}}{{\vect{{u}}}}={\vect{\hat{f}}}
  \label{eqn:NavierStokesequationALE2general}
\end{equation}
where ${\vect{\hat{u}}}$ is the vector of unknown ``DOFs'', $\matr{K}$ is the
stiffness matrix, $\fnof{\vect{\hat{g}}}{{\vect{{u}}}}$ a non-linear vector
function and ${\vect{\hat{f}}}$ the forcing vector. In
\eqnref{eqn:NavierStokesweakform} the only real non-linear term is represented
by
$\fnof{\vect{\hat{g}}}{\vect{{u}}}=\gint{\Omega}{}{\rho\pbrac{\dotprod{\vectr{u}}{\grad}}\vectr{u}\vectr{v}
}{\Omega}$.  If $\fnof{\vect{\hat{g}}}{\vect{u}}$ is not $\equiv\vect{0}$ then
we use Newton's method \ie
\begin{equation}
  \begin{split}
    \text{1.  } & \fnof{\matr{J}}{\vect{u}_{i}}.\delta
    \vect{u}_{i} = 
    -\fnof{\vect{\psi}}{\vectr{u}_{i}} \\
    \text{2.  } & \vect{u}_{i+1}=\vect{u}_{i}+\delta
    \vect{u}_{i}
  \end{split}
\end{equation}
where $\fnof{\matr{J}}{\vect{u}}$ is the Jacobian and is given by
\begin{equation}
  \fnof{\matr{J}}{\vect{u}}=\matr{K}+
    \delby{\fnof{\vect{\hat{g}}}{\vect{u}}}{\vect{u}}
\end{equation}
with the stiffness matrix $\matr{K}$ derived from
\eqnref{eqn:NavierStokesequationALE2general} by applying Green's theorem as
follows:
\begin{equation}
  \begin{split}
  \matr{K}\vectr{\hat{u}}=
  \gint{\Omega}{}{\divergence{}{\vectr{v}}p}{\Omega}
  -\gint{\Omega}{}{\doubledotprod{\mu\gradient{}{\vectr{v}}}{\gradient{}{\vectr{u}}}}{\Omega}
  -\gint{\Omega}{}{\rho\pbrac{\dotprod{\vectr{u}^*}{\grad}}\vectr{u}\vectr{v}}{\Omega}
  +\gint{\Omega}{}{\divergence{}{\vectr{u}}q}{\Omega}
  \end{split}
  \label{eqn:NavierStokesweakform2}
\end{equation}
and $\fnof{\vect{\psi}}{\vect{\hat{u}}}=\matr{K}\vect{\hat{u}}+\fnof{\vect{\hat{g}}}{\vect{u}}+\vect{\hat{f}}$.


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../../OpenCMISSNotes"
%%% End: 
